---
title: "CDEnrich"
output:
  github_document:
    toc: true
    toc_depth: 2
editor_options:
  markdown:
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd instead. -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
BiocParallel::register(BiocParallel::SerialParam())
```

# CDEnrich

<!-- badges: start -->

<!-- badges: end -->

> Curated gene sets for 27 cell death modalities + one-stop enrichment
> analysis (ORA / GSEA / ssGSEA) with publication-ready visualization.

## Overview

`CDEnrich` provides manually curated gene sets for **27 regulated cell
death modalities** — 13 PMID-supported (Ferroptosis, Cuproptosis,
Disulfidptosis, Apoptosis, Pyroptosis, Autophagy, Necroptosis, NETosis,
Immunogenic Cell Death, Anoikis, Phagocytosis, Efferocytosis,
PANoptosis) and 14 GeneCards-based (exploratory). It supports ORA, GSEA,
ssGSEA, differential ssGSEA, multi-group comparison, and visualization
through a unified interface.

## Installation

```{r, eval=FALSE}
# Development version
remotes::install_github("Yushan-Chen-829/CDEnrich")

# After Bioconductor acceptance:
# BiocManager::install("CDEnrich")
```

Bioconductor dependencies (`clusterProfiler`, `enrichplot`, `fgsea`,
`GSVA`) are installed automatically; install `BiocManager` first if
needed.

## Quick start

```{r, eval=FALSE}
library(CDEnrich)

# Load all curated gene sets
gene_sets <- load_cell_death_genes()

# Recommended representative pathway (with PMID & source)
get_representative_pathway("Ferroptosis")
```

### ORA

```{r, eval=FALSE}
data("demo_deg", package = "CDEnrich")
data("demo_all_genes", package = "CDEnrich")

# Global (all gene sets) or targeted (recommended set)
ora_all  <- celldeath_enrich(deg = demo_deg, universe = demo_all_genes)
ora_ferr <- celldeath_enrich(deg = demo_deg, universe = demo_all_genes,
                             term_select = "Ferroptosis")
```

### GSEA

```{r, eval=FALSE}
data("demo_rank_apop", package = "CDEnrich")
gsea_result <- celldeath_gsea(geneList = demo_rank_apop, pvalueCutoff = 0.05)
```

### ssGSEA

```{r, eval=FALSE}
data("demo_expr", package = "CDEnrich")
data("demo_group", package = "CDEnrich")

ssgsea_result <- celldeath_ssgsea(expr = demo_expr, group = demo_group)
ssgsea_diff   <- celldeath_ssgsea_diff(ssgsea_result)  # Wilcoxon / Kruskal-Wallis
```

### Multi-group

```{r, eval=FALSE}
data("demo_deg_list", package = "CDEnrich")
data("demo_rank_apop_list", package = "CDEnrich")

compare_ora <- celldeath_compare_enrich(deg_list_list = demo_deg_list,
                                        universe = demo_all_genes)
multi_gsea  <- celldeath_gsea_multiple(geneList_list = demo_rank_apop_list)
```

## Selecting pathways

The `term_select` argument works identically across all analysis
functions:

| Usage | Syntax |
|----|----|
| All gene sets | Omit `term_select` |
| Recommended set for a death type | `term_select = "Ferroptosis"` |
| Exact gene set name | `term_select = "Ferroptosis_FerrDb V2(36305834)"` |
| All sets under one death type | Add `use_recommended = FALSE` |

## Visualization

```{r, eval=FALSE}
plot_death_enrich_bubble(ora_all)
plot_death_enrich_targeted(ora_ferr)
plot_death_gsea_curve(gsea_result, pathway = "Ferroptosis")
plot_death_ssgsea_heatmap(ssgsea_result)
```

| Category | Functions |
|----|----|
| ORA | `plot_death_enrich_bubble()`, `plot_death_enrich_network()`, `plot_death_enrich_targeted()` |
| Compare | `plot_death_compare_bubble()`, `plot_death_compare_heatmap()` |
| GSEA | `plot_death_gsea_curve()`, `plot_death_gsea_curve_multiple()`, `plot_death_gsea_heatmap_multiple()`, `plot_death_gsea_ranked()` |
| ssGSEA | `plot_death_ssgsea_bar()`, `plot_death_ssgsea_boxplot()`, `plot_death_ssgsea_heatmap()` |
| Differential | `plot_death_volcano()` |

Most functions support `savefile` / `filename` / `filename_prefix` to
export results directly.

## Citation

Please cite `CDEnrich` and the original publications of the curated gene
sets. Retrieve PMID and source information with
`get_representative_pathway("Ferroptosis")`.

## License

MIT License. **Yushan Chen** —
[chen_yushan0625\@163.com](mailto:chen_yushan0625@163.com) ·
[Issues](https://github.com/Yushan-Chen-829/CDEnrich/issues)
