---
title: "CDEnrich"
output:
  github_document:
    toc: true
    toc_depth: 2
editor_options:
  markdown:
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd instead. -->

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
BiocParallel::register(BiocParallel::SerialParam())
```

# CDEnrich

<!-- badges: start -->
<!-- badges: end -->

## Overview

`CDEnrich` provides curated gene sets for multiple cell death types
(Ferroptosis, Cuproptosis, Disulfidptosis, Apoptosis, Pyroptosis,
Autophagy, Necroptosis, NETosis, Immunogenic Cell Death, Anoikis,
Phagocytosis, Efferocytosis, and PANoptosis) and supports:

- ORA, GSEA, ssGSEA, and differential ssGSEA analysis
- Multi-group enrichment comparison
- Bubble plots, heatmaps, GSEA curves, barplots, boxplots, and volcano plots
- Recommended representative gene sets with PMID information

## Installation

```{r, eval=FALSE}
install.packages("remotes")

# Bioconductor dependencies
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("clusterProfiler", "enrichplot", "fgsea", "GSVA"))

# Install from GitHub (replace with your repository once published)
remotes::install_github("Yushan-Chen-829/CDEnrich")
```

All remaining CRAN dependencies are installed automatically when declared
in `DESCRIPTION`.

## Quick start

```{r, eval=FALSE}
library(CDEnrich)

# Load all curated gene sets
gene_sets <- load_cell_death_genes()

# Recommended representative gene set for a cell death type
# (returns PMID, source, and citation information)
get_representative_pathway("Ferroptosis")
```

## ORA

```{r, eval=FALSE}
genes <- c("GPX4", "ACSL4", "SLC7A11", "FTH1", "TFRC",
           "GSDMD", "CASP1", "CASP3", "BAX")

# Global ORA: test against all curated gene sets
ora_result <- celldeath_enrich(deg = genes)

# Targeted ORA: uses the recommended representative gene set by default
ora_ferroptosis <- celldeath_enrich(deg = genes, term_select = "Ferroptosis")
```

For formal analysis, provide an appropriate background gene set via the
`universe` argument.

## GSEA

The input is a named numeric vector sorted in decreasing order:

```{r, eval=FALSE}
gene_rank <- c(GPX4 = 2.8, ACSL4 = 2.4, SLC7A11 = 2.1, FTH1 = 1.7,
               TFRC = 1.4, GSDMD = 0.8, CASP1 = 0.4, CASP3 = -1.2, BAX = -1.8)
gene_rank <- sort(gene_rank, decreasing = TRUE)

gsea_result <- celldeath_gsea(geneList = gene_rank)
```

## ssGSEA

`expr` is an expression matrix with genes in rows and samples in columns:

```{r, eval=FALSE}
ssgsea_result <- celldeath_ssgsea(expr = expr, group = group)
ssgsea_diff   <- celldeath_ssgsea_diff(ssgsea_result)
```

## Multi-group analysis

```{r, eval=FALSE}
# Multi-group ORA: a named list of gene vectors
compare_result <- celldeath_compare_enrich(deg_list_list = gene_groups)

# Multi-group GSEA: a named list of ranked gene vectors
gsea_groups <- celldeath_gsea_multiple(geneList_list = ranked_groups)
```

## Selecting pathways

The `term_select` argument follows the same rules in all analysis
functions:

| Usage | Syntax |
| --- | --- |
| All gene sets (global mode) | Omit `term_select` |
| Recommended gene set for a cell death type | `term_select = "Ferroptosis"` |
| Exact gene set name (case-sensitive) | `term_select = "Disulfidptosis_Up_36747082"` |
| All gene sets under a cell death type | Add `use_recommended = FALSE` |

## Visualization

```{r, eval=FALSE}
plot_death_enrich_bubble(ora_result)
plot_death_gsea_curve(gsea_result, pathway = "Ferroptosis")
plot_death_ssgsea_heatmap(ssgsea_result)
```

All plotting functions: `plot_death_enrich_bubble()`,
`plot_death_enrich_network()`, `plot_death_compare_bubble()`,
`plot_death_compare_heatmap()`, `plot_death_gsea_curve()`,
`plot_death_gsea_curve_multiple()`, `plot_death_gsea_heatmap_multiple()`,
`plot_death_gsea_ranked()`, `plot_death_ssgsea_bar()`,
`plot_death_ssgsea_boxplot()`, `plot_death_ssgsea_heatmap()`, and
`plot_death_volcano()`.

## Saving results

Most analysis and plotting functions can save output files through the
`savefile`, `filename`, or `filename_prefix` arguments:

```{r, eval=FALSE}
ora_result <- celldeath_enrich(deg = genes, savefile = TRUE,
                               filename = "cell_death_ora.csv")
plot_death_enrich_bubble(ora_result, filename = "cell_death_bubble.png")
```

## Development

```{r, eval=FALSE}
devtools::document()      # Regenerate documentation
devtools::test()          # Run the test suite
devtools::build_readme()  # Build README.md from README.Rmd
devtools::check()         # Build and check the package
```

## Citation

If you use `CDEnrich` in a research project, please cite the package and
the original publications of the curated gene sets. PMID and source
information for recommended representative pathways can be obtained with
`get_representative_pathway("Ferroptosis")`.

## License

`CDEnrich` is licensed under the MIT License.

## Contact

Yushan Chen — [chen_yushan0625@163.com](mailto:chen_yushan0625@163.com)
