---
title: "CDEnrich: Complete Usage Guide for Cell Death Enrichment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{CDEnrich: Complete Usage Guide for Cell Death Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 8,
  fig.height = 5,
  dpi = 300
)
```

# 1. Introduction

CDEnrich is a dedicated, one-stop enrichment analysis toolkit for
programmed cell death pathway research. It focuses specifically on cell
death-related enrichment analysis and publication-grade visualization,
covering three mainstream enrichment algorithms: ORA
(Over-Representation Analysis), GSEA (Gene Set Enrichment Analysis) and
ssGSEA (single-sample GSEA).

This package integrates manually curated, high-quality gene sets for
multiple programmed cell death patterns, including apoptosis,
ferroptosis, pyroptosis, necroptosis, and cuproptosis. It supports both
single-group and multi-group comparative enrichment analysis, with a
complete set of matched visualization functions for all analysis
scenarios.

**Key Features**

-   Exclusive standardized gene sets for programmed cell death pathways
-   Supports single-group and multi-group ORA enrichment analysis
-   Supports single-group and multi-group GSEA enrichment analysis
-   Supports ssGSEA-based sample-level scoring and multi-group
    differential analysis - Scenario-matched visualization functions for
    all analysis modes
-   All plots are fully customizable and ready forpublication

# 2. Installation

CDEnrich depends on several Bioconductor packages (clusterProfiler,
GSVA, enrichplot). Install them first if you do not have them:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("clusterProfiler", "GSVA", "enrichplot"))
```

Then install the development version of CDEnrich from GitHub:

```{r install-github,eval=FALSE}
# Install the remotes package if needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

# Install CDEnrich (replace "YourUserName" with your GitHub account)
remotes::install_github("Yushan-Chen-829/CDEnrich", build_vignettes = TRUE)
```

Load the package and verify the installation:

```{r load-pkg}
library(CDEnrich)
packageVersion("CDEnrich")
```

# 3. Built-in Cell Death Gene Sets

CDEnrich provides a comprehensive collection of 27 cell death-related
gene sets as the reference database for enrichment analysis. These gene
sets were manually curated from two sources:

-   **PMID + GeneCards** (19 gene sets): Curated by integrating
    PubMed-indexed literature with GeneCards annotations. These sets are
    considered high-confidence and include major cell death types such
    as *Apoptosis*, *Ferroptosis*, *Pyroptosis*, *Autophagy*,
    *Necroptosis*, and others.

-   **GeneCards only** (8 gene sets): Curated solely from GeneCards
    annotations without explicit PMID validation in the current version,
    including *Parthanatos*, *Entotic cell death*, *Alkaliptosis*,
    *Oxeiptosis*, *Paraptosis*, *Methuosis*, and *paraptosis‑like
    death*.

CDEnrich ships curated gene sets covering 27 cell death types, named
with the convention `<Type>_<Source>_<PMID>` (e.g.
`Ferroptosis_FerrDb V2(36305834)`).\
You can inspect available cell death pathway names with the following
command:

```{r load-data}
gene_sets <- load_cell_death_genes()
length(gene_sets)      # number of gene sets
names(gene_sets)       # all available gene set names
```

## 3.1 Recommended representative gene sets

Many cell death types have MULTIPLE gene sets curated from different
sources and publications. To help users choose, CDEnrich designates a
recommended representative gene set for 13 cell death types:

```{r recommendation-genesets}
# Recommendation metadata: type, pathway name, PMID, source
representative_pathways

# Look up the recommendation for one type
get_representative_pathway("Disulfidptosis")

# Types without a recommendation return NULL
get_representative_pathway("Mitosis")
```

## 3.2 Three ways to select pathways(`term_select`)

All analysis functions share the same `term_select` logic:

1.  `NULL` (default): run against ALL cell death gene sets (global
    mode);

2.  A cell death type name (e.g. `"Ferroptosis"`): the recommended
    representative set is used automatically, with an informative
    message; set `use_recommended = FALSE` to run against ALL sets of
    that type;

3.  An exact gene set name (e.g. `"Ferroptosis_FerrDb V2(36305834)"`):
    used directly, bypassing the recommendation.

```{r term-select-demo}
# Demo DEG list (see Section 6.1 for details)
deg <- c("GPX4", "ACSL4", "SLC7A11", "GSDMD", "CASP1", "CASP3", "BAX")

# Recommended set is picked automatically (watch the message!)
celldeath_enrich(deg, term_select = "Ferroptosis")

# Run against ALL Ferroptosis gene sets instead
celldeath_enrich(deg, term_select = "Ferroptosis", use_recommended = FALSE)
```

# 4. Workflow Overview

CDEnrich provides four core analysis scenarios with dedicated supporting
visualization functions. The correspondence between analysis modes and
plotting tools is summarized below.

## Supported Analyses

-   **Over-Representation Analysis (ORA)**: Identifies over-represented
    cell death pathways in a given gene list.
-   **Gene Set Enrichment Analysis (GSEA)**: Evaluates the distribution
    of cell death gene sets across a ranked list of all genes.
-   **Single-Sample GSEA (ssGSEA)**: Computes a cell death pathway
    activity score for each sample, enabling sample- and group-level
    comparison.

## Supported Visualization

### Single-group ORA Enrichment Visualization

-   Bubble plot: `plot_death_enrich_bubble()`
-   Pathway enrichment network plot: `plot_death_enrich_network()`
-   Enrichment volcano plot: `plot_death_volcano()`

### Single-group GSEA Enrichment Visualization

-   Standard GSEA curve plot: `plot_death_gsea_curve()`
-   GSEA running ranking score plot: `plot_death_gsea_ranked()`
-   Enrichment volcano plot: `plot_death_volcano()`

### Multi-group Comparative ORA Visualization

-   Comparative bubble plot: `plot_death_compare_bubble()`
-   Comparative enrichment heatmap: `plot_death_compare_heatmap()`

### Multi-group Comparative GSEA Visualization

-   Batch GSEA curve plot across groups:
    `plot_death_gsea_curve_multiple()`
-   Multi-group GSEA statistic heatmap:
    `plot_death_gsea_heatmap_multiple()`

### ssGSEA Visualization

-   Score heatmap with group annotation: `plot_death_ssgsea_heatmap()`
-   Grouped boxplot: `plot_death_ssgsea_boxplot()`
-   Mean score barplot: `plot_death_ssgsea_bar()`

# 5. Input Data Specification

All analytical functions in CDEnrich require strictly formatted inputs.
We summarize required input formats for all four analysis workflows:

## Single-group ORA (`celldeath_enrich()`)

-   Input: Character vector of gene symbols (DE genes).
-   Minimum length: \>=3 genes.

## Single-group GSEA (`celldeath_gsea()`)

-   Input: Named numeric vector, **pre-sorted in descending order**.
-   Names: Gene symbols; Values: ranking metric (log2FC / t statistic).

## Multi-group Comparative ORA (`celldeath_compare_enrich()`)

-   Input: Named list. Each element = character vector of DE genes for
    one group.
-   List names = group labels. Each gene vector requires \>=3 genes.

## Multi-group Comparative GSEA (`celldeath_gsea_multiple()`)

-   Input: Named list. Each element = named numeric vector sorted
    descending.
-   List names = group labels.

## ssGSEA (`celldeath_ssgsea()`)

-   Input: Numeric matrix or data.frame, genes (symbols) as ROWS x
    samples as COLUMNS.
-   Rownames/colnames required; no NA, no duplicated gene symbols.
-   Optional `group`: group label vector, length = number of samples
    (\>= 2 groups, \>= 2 samples per group recommended).

# 6. Minimal Working Examples

## 6.1 Single-group ORA Enrichment Example

```{r single-ORA, eval=FALSE}
library(CDEnrich)

# Load the built-in demonstration data
data("demo_deg")
data("demo_all_genes")

# Input: character vector of differentially expressed gene symbols
demo_deg

# Global mode: enrichment against all cell death pathways
ora_res <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes
)

# Visualization
plot_death_enrich_bubble(ora_res)
plot_death_enrich_network(ora_res)
plot_death_volcano(ora_res)

# Targeted mode: enrich against the recommended Ferroptosis gene set
ora_single <- celldeath_enrich(
  deg = demo_deg,
  term_select = "Ferroptosis",
  universe = demo_all_genes
)

# Inspect targeted enrichment result
ora_single

# Visualization for targeted mode

# Targeted mode returns only one pathway, so multi-pathway plots
# (bubble / network / volcano) are not informative here.
# Use the dedicated compact summary plot instead.
plot_death_enrich_targeted(ora_single)

# Note: plot_death_enrich_bubble(), plot_death_enrich_network() and
# plot_death_volcano() also accept targeted results — they detect the
# single-pathway mode and automatically switch to the same compact
# summary plot shown above.

# Export targeted results as CSV
ora_single <- celldeath_enrich(
  deg = demo_deg,
  term_select = "Ferroptosis",
  universe = demo_all_genes,
  savefile = TRUE,
  filename = "single_group_Ferroptosis_ORA_result.csv"
)
```

## 6.2 Single-group GSEA Enrichment Example

```{r single-GSEA}
library(CDEnrich)

# Load the built-in Apoptosis demonstration ranked list
data("demo_rank_apop")

# The representative Apoptosis gene set (PMID 37968457) contains
# apoptosis-related genes collected from MSigDB, KEGG, GeneCards
# and relevant reviews. These genes are intentionally placed at the
# top of the ranked list to create a clear enrichment signal.
apoptosis_id <- get_representative_pathway("Apoptosis")$Pathway

# Global mode: GSEA against all cell death pathways
gsea_res <- celldeath_gsea(
  geneList = demo_rank_apop,
  pvalueCutoff = 0.05
)
```

Note: fgsea may report warnings about very small p-values for this
demonstration because the Apoptosis signal is intentionally strong.
These warnings do not affect the enrichment results.

```{r}
# Visualization for global mode
plot_death_gsea_curve(gsea_res, pathway = apoptosis_id)
plot_death_gsea_ranked(gsea_res, pathway = apoptosis_id)
plot_death_volcano(gsea_res)

# Targeted mode: test only the Apoptosis gene set
gsea_single <- celldeath_gsea(
  geneList = demo_rank_apop,
  term_select = apoptosis_id,
  pvalueCutoff = 0.05
)

# Inspect targeted enrichment result
gsea_single

# Visualization for targeted mode
plot_death_gsea_curve(gsea_single, pathway = apoptosis_id)
plot_death_gsea_ranked(gsea_single, pathway = apoptosis_id)

# Note: plot_death_volcano() also accepts gsea_single, but targeted GSEA
# contains only one pathway and therefore produces a one-point plot.
```

```{r eval=FALSE}
# Export global GSEA results as CSV
celldeath_gsea(
  geneList = demo_rank_apop,
  pvalueCutoff = 0.05,
  savefile = TRUE,
  filename = "single_group_GSEA_result.csv"
)
```

## 6.3 Multi-group Comparative ORA Enrichment Examples

```{r multi-ORA, eval=FALSE}
library(CDEnrich)

# Load the built-in demonstration data
data("demo_deg_list")
data("demo_all_genes")

# Inspect the named list of DE gene symbols
demo_deg_list

# Global mode: compare all groups against all cell death pathways
compare_ora_global <- celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  universe = demo_all_genes,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize = 3
)

# Visualization for global multi-group ORA results
plot_death_compare_bubble(compare_ora_global)
plot_death_compare_heatmap(compare_ora_global)

# Targeted mode: compare all groups against Ferroptosis
compare_ora_targeted <- celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  term_select = "Ferroptosis",
  universe = demo_all_genes,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize = 3
)

# Inspect targeted multi-group ORA results
compare_ora_targeted

# Targeted results contain one pathway across multiple groups.
# Therefore, the comparison bubble plot and heatmap remain informative.
plot_death_compare_bubble(compare_ora_targeted)
plot_death_compare_heatmap(compare_ora_targeted)
```

```{r eval=FALSE}
# Export global multi-group ORA results as CSV
celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  universe = demo_all_genes,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize = 3,
  savefile = TRUE,
  filename = "multi_group_ORA_result.csv"
)

# Export targeted multi-group ORA results as CSV
celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  term_select = "Ferroptosis",
  universe = demo_all_genes,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize = 3,
  savefile = TRUE,
  filename = "multi_group_Ferroptosis_ORA_result.csv"
)
```

## 6.4 Multi-group Comparative GSEA Example

```{r multi-GSEA, eval=FALSE}
library(CDEnrich)

# Load the built-in multi-group demonstration ranked lists
data("demo_rank_apop_list")

# Input requirement: a NAMED list, each element is a named numeric
# vector sorted in descending order.
# The Control group has no introduced signal; the Treatment group
# carries a strong Apoptosis signal.
str(demo_rank_apop_list)

apoptosis_id <- get_representative_pathway("Apoptosis")$Pathway

# Global mode: compare all groups against all cell death pathways
multi_gsea_global <- celldeath_gsea_multiple(
  geneList_list = demo_rank_apop_list,
  pvalueCutoff = 1
)

#Note: pvalueCutoff is set to 1 in multi-group mode so that every group retains a complete result table. This keeps the cross-group comparison meaningful — groups without a real signal (Control) will show small NES and large p-values rather than disappearing from the plots.

# Visualization

# Batch plot GSEA curves across all groups for the selected pathway
curves <- plot_death_gsea_curve_multiple(multi_gsea_global, pathway = apoptosis_id)
curves$Control
curves$Treatment

# Multi-group GSEA statistics heatmap
plot_death_gsea_heatmap_multiple(multi_gsea_global)

# Targeted mode: all groups compared against the Apoptosis gene set
multi_gsea_targeted <- celldeath_gsea_multiple(
  geneList_list = demo_rank_apop_list,
  term_select = apoptosis_id,
  pvalueCutoff = 1
)

# Inspect targeted multi-group results
multi_gsea_targeted

# Targeted visualization: curves per group remain informative
plot_death_gsea_curve_multiple(multi_gsea_targeted, pathway = apoptosis_id)
```

```{r eval=FALSE}
# Export each group's GSEA result as separate CSV files
celldeath_gsea_multiple(
  geneList_list = demo_rank_apop_list,
  pvalueCutoff = 0.05,
  savefile = TRUE
)
```

## 6.5 ssGSEA and Multi-group Differential Analysis

Unlike ORA/GSEA, ssGSEA scores EVERY SAMPLE on EVERY pathway, so no
prior differential analysis is needed. All samples are scored together
on the same expression background to guarantee score comparability
across groups.

```{r ssGSEA}
library(CDEnrich)

# Load the built-in demonstration expression data
data("demo_expr")
data("demo_group")

# expr: genes in rows, samples in columns
# demo_group: 6 Control and 6 Treatment samples
dim(demo_expr)
demo_group

# Global mode with group labels
ssgsea_res <- celldeath_ssgsea(demo_expr, group = demo_group)
ssgsea_res   # print method: quick summary of pathways / samples / groups

# Targeted mode (uses the recommended representative set)
ssgsea_ferro <- celldeath_ssgsea(
  demo_expr,
  group = demo_group,
  term_select = "Ferroptosis"
)

# Differential analysis between groups
# (Wilcoxon for 2 groups, Kruskal-Wallis for >= 3 groups)
diff_res <- celldeath_ssgsea_diff(ssgsea_res)
head(diff_res)

# Visualization
plot_death_ssgsea_heatmap(ssgsea_res)   # top pathways, z-scored, group annotation
plot_death_ssgsea_boxplot(ssgsea_res)   # per-pathway group comparison
plot_death_ssgsea_bar(ssgsea_res)       # mean score per group with SEM
```

```{r eval=FALSE}
# Export ssGSEA score matrix as CSV
celldeath_ssgsea(
  demo_expr,
  group = demo_group,
  savefile = TRUE,
  filename = "single_group_ssGSEA_score.csv"
)

# Export differential analysis results as CSV
celldeath_ssgsea_diff(
  ssgsea_res,
  savefile = TRUE,
  filename = "single_group_ssGSEA_diff.csv"
)
```

# 7. Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
