---
title: "CDEnrich: Complete Usage Guide for Cell Death Enrichment"
author:
  - name: Yushan Chen
    email: chen_yushan0625@163.com
output:
  BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{CDEnrich: Complete Usage Guide for Cell Death Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 5,
  dpi = 300
)

library(CDEnrich)
```

# Introduction

CDEnrich is an R package for enrichment analysis of curated cell death
pathways. It provides gene sets for multiple forms of regulated cell
death, including apoptosis, ferroptosis, pyroptosis, necroptosis,
cuproptosis, and disulfidptosis.

The package supports over-representation analysis (ORA), gene set
enrichment analysis (GSEA), single-sample GSEA (ssGSEA), multi-group
comparison, and visualization of enrichment results.

This vignette demonstrates a complete workflow using built-in
demonstration data.

**Key Features**

-   Exclusive standardized gene sets for programmed cell death pathways
-   Supports single-group and multi-group ORA enrichment analysis
-   Supports single-group and multi-group GSEA enrichment analysis
-   Supports ssGSEA-based sample-level scoring and multi-group
    differential analysis - Scenario-matched visualization functions for
    all analysis modes
-   All plots are fully customizable and ready forpublication

# Installation

`CDEnrich` and its dependencies should be installed before running this
vignette. The current development version can be installed from GitHub:

```{r install-github, eval=FALSE}
remotes::install_github("Yushan-Chen-829/CDEnrich")
```

After acceptance into Bioconductor, the package will be available
through `BiocManager`:

```{r install-bioconductor, eval=FALSE}
BiocManager::install("CDEnrich")
```

Load the package as follows:

```{r load-pkg}
library(CDEnrich)
packageVersion("CDEnrich")
```

# Cell Death Gene Sets

CDEnrich provides curated gene sets representing 27 cell death types.
The gene sets were compiled from published literature, PMID-supported
annotations, and GeneCards.

The 27 cell death types are organized into two source categories:

-   **PMID-supported literature and GeneCards (13 types)**:
    Disulfidptosis, Ferroptosis, Cuproptosis, Apoptosis, Pyroptosis,
    Autophagy, Necroptosis, NETosis, Immunogenic Cell Death, Anoikis,
    Phagocytosis, Efferocytosis, and PANoptosis.

-   **GeneCards (14 types)**: Parthanatos, Entotic cell death,
    Lysosome-dependent cell death, Alkaliptosis, Oxeiptosis, Paraptosis,
    Methuosis, Intrinsic apoptosis, Extrinsic apoptosis, Mitochondrial
    permeability transition (MPT)-driven necrosis, Autosis (autophagic
    cell death), Mitotic death, Accidental cell death, and
    paraptosis-like death.

GeneCards-based gene sets should be interpreted as exploratory
resources, particularly when no PMID is available in the package
metadata.

## Available Gene Sets

The curated gene sets can be loaded with:

```{r load-gene-sets}
gene_sets <- load_cell_death_genes()

length(gene_sets)
head(names(gene_sets))
```

The returned object is a named list in which each element is a character
vector of gene symbols. The number of gene sets may differ from 27
because one cell death type can contain multiple gene sets.

The size of each gene set can be inspected with:

```{r gene-set-summary}
gene_set_summary <- data.frame(
  gene_set = names(gene_sets),
  n_genes = lengths(gene_sets),
  row.names = NULL
)

head(gene_set_summary)
```

Gene set names may include the cell death type, source, and PMID. For
example:

``` r
"Ferroptosis_FerrDb V2(36305834)"
```

## Representative Gene Sets

For selected cell death types, `CDEnrich` provides a recommended
representative gene set:

```{r representative-pathway}
get_representative_pathway("Ferroptosis")
get_representative_pathway("Disulfidptosis")
```

The returned information may include the cell death type, pathway name,
PMID, source, and description. If no representative pathway is
available, the function returns `NULL`.

```{r representative-pathway-null}
get_representative_pathway("Mitosis")
```

## Pathway Selection

All enrichment functions share the same `term_select` logic:

### Global analysis

Omit `term_select` to use all curated gene sets:

```{r global-selection}
data("demo_deg", package = "CDEnrich")
data("demo_all_genes", package = "CDEnrich")

ora_global <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes
)
```

### Recommended representative pathway

Provide a cell death type to use its recommended gene set:

```{r recommended-selection}
ora_ferroptosis <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes,
  term_select = "Ferroptosis"
)
```

### All gene sets within a cell death type

Set `use_recommended = FALSE` to use all gene sets associated with a
cell death type:

```{r all-selection}
ora_ferroptosis_all <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes,
  term_select = "Ferroptosis",
  use_recommended = FALSE
)
```

### Exact gene set name

An exact gene set name can be supplied. The name must match the output
of `load_cell_death_genes()` exactly:

```{r exact-selection}
ora_exact <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes,
  term_select = "Ferroptosis_FerrDb V2(36305834)"
)
```

The same pathway-selection rules apply to GSEA, ssGSEA, and multi-group
analysis.

# Workflow Overview

`CDEnrich` provides three main enrichment workflows:

-   **Over-Representation Analysis (ORA):** Identifies cell death
    pathways over-represented in a gene list.
-   **Gene Set Enrichment Analysis (GSEA):** Evaluates cell death
    pathways across a ranked gene list.
-   **Single-Sample GSEA (ssGSEA):** Calculates a cell death pathway
    score for each sample and supports group-level comparisons.

The package also supports multi-group ORA and GSEA, differential
analysis of ssGSEA scores, and visualization of enrichment results.

The general workflow is:

``` text
Input data
  -> Select cell death gene sets
  -> Run ORA, GSEA, or ssGSEA
  -> Compare groups when appropriate
  -> Visualize and interpret the results
```

The main analysis functions are:

| Analysis            | Function                     |
|---------------------|------------------------------|
| Single-group ORA    | `celldeath_enrich()`         |
| Multi-group ORA     | `celldeath_compare_enrich()` |
| Single-group GSEA   | `celldeath_gsea()`           |
| Multi-group GSEA    | `celldeath_gsea_multiple()`  |
| ssGSEA              | `celldeath_ssgsea()`         |
| Differential ssGSEA | `celldeath_ssgsea_diff()`    |

# Input Data Specification

The main input formats required by each analysis function are summarized
below.

## Single-group ORA

`celldeath_enrich()` accepts a character vector of gene symbols.

**Input requirements:**

-   `deg`: A character vector of gene symbols.
-   At least three genes are required.
-   `universe`: An optional character vector of background genes.

```{r ora-input}
ora_genes <- c("GPX4", "ACSL4", "SLC7A11")
```

## Single-group GSEA

`celldeath_gsea()` requires a named numeric vector sorted in decreasing
order.

**Input requirements:**

-   Names: Gene symbols.
-   Values: A ranking statistic, such as log2 fold change or a test
    statistic.
-   Order: Values must be sorted in decreasing order.

```{r gsea-input}
gene_rank <- c(
  GPX4 = 2.8,
  ACSL4 = 2.4,
  SLC7A11 = 2.1
)

gene_rank <- sort(gene_rank, decreasing = TRUE)
```

## Multi-group ORA

`celldeath_compare_enrich()` accepts a named list of gene vectors.

**Input requirements:**

-   Each list element must be a character vector of gene symbols.
-   The list must be named.
-   List names represent the group labels.
-   Each gene vector must contain at least three genes.

```{r multi-ora-input}
gene_groups <- list(
  Control = c("GPX4", "ACSL4", "FTH1"),
  Treatment = c("GPX4", "SLC7A11", "GSDMD")
)
```

## Multi-group GSEA

`celldeath_gsea_multiple()` accepts a named list of ranked gene vectors.

**Input requirements:**

-   Each list element must be a named numeric vector.
-   Each vector must be sorted in decreasing order.
-   The list must be named.
-   List names represent the group labels.

```{r multi-gsea-input}
ranked_groups <- list(
  Control = c(GPX4 = 2.1, ACSL4 = 1.5, SLC7A11 = 0.8),
  Treatment = c(GPX4 = 2.8, ACSL4 = 2.4, SLC7A11 = 1.9)
)
```

## ssGSEA

`celldeath_ssgsea()` accepts a numeric matrix or data frame with genes
in rows and samples in columns.

**Input requirements:**

-   Rows represent genes; columns represent samples.
-   Row names must contain gene symbols.
-   Column names must contain sample names.
-   Missing values are not allowed.
-   Duplicated gene symbols are not allowed.
-   The optional `group` vector must contain one label for each sample.

```{r ssgsea-input}
expr <- matrix(
  rnorm(30),
  nrow = 3,
  ncol = 10,
  dimnames = list(
    c("GPX4", "ACSL4", "SLC7A11"),
    paste0("Sample", 1:10)
  )
)

group <- rep(c("Control", "Treatment"), each = 5)
```

# Minimal Working Examples

## Single-group ORA

The following example performs global ORA using all curated gene sets:

```{r single-group-ora}
data("demo_deg", package = "CDEnrich")
data("demo_all_genes", package = "CDEnrich")

ora_global <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes
)

ora_global
```

A bubble plot can be used to visualize the enriched pathways:

```{r single-group-ora-plot}
plot_death_enrich_bubble(
  ora_global,
  show_category = 10
)
```

A targeted analysis can be performed by specifying a cell death type.
The recommended representative gene set is selected automatically:

```{r targeted-ora}
ora_ferroptosis <- celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes,
  term_select = "Ferroptosis"
)

ora_ferroptosis
```

The `universe` argument should represent the genes that were available
for detection or testing in the original experiment.

The targeted enrichment result can be visualized with
`plot_death_enrich_targeted()`:

```{r targeted-ora-plot}
plot_death_enrich_targeted(
  ora_ferroptosis
)
```

This plot provides a compact visualization for a targeted analysis
containing a single representative pathway.

## Single-group GSEA

The following example performs GSEA using a built-in ranked gene list:

```{r single-group-gsea}
data("demo_rank_apop", package = "CDEnrich")

apoptosis_id <- get_representative_pathway("Apoptosis")$Pathway

gsea_global <- celldeath_gsea(
  geneList = demo_rank_apop,
  pvalueCutoff = 0.05
)

gsea_global
```

The enrichment profile of the representative apoptosis pathway can be
visualized with:

```{r single-group-gsea-plot}
plot_death_gsea_curve(
  gsea_global,
  pathway = apoptosis_id
)
```

A targeted GSEA can be performed using the representative apoptosis gene
set:

```{r targeted-gsea}
gsea_apoptosis <- celldeath_gsea(
  geneList = demo_rank_apop,
  term_select = apoptosis_id,
  pvalueCutoff = 0.05
)

gsea_apoptosis
```

The demonstration ranking list contains a strong apoptosis signal.
Depending on the installed version of `fgsea`, very small p-values may
be reported.

## Multi-group ORA

Multi-group ORA requires a named list of gene vectors:

```{r multi-group-ora}
data("demo_deg_list", package = "CDEnrich")
data("demo_all_genes", package = "CDEnrich")

compare_ora <- celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  universe = demo_all_genes,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize = 3
)

compare_ora
```

The results can be visualized with a comparative heatmap:

```{r multi-group-ora-plot}
plot_death_compare_heatmap(
  compare_ora,
  show_category = 10
)
```

A targeted comparison can be performed by specifying a cell death type:

```{r targeted-multi-group-ora}
compare_ora_ferroptosis <- celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  universe = demo_all_genes,
  term_select = "Ferroptosis",
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2,
  minGSSize = 3
)

compare_ora_ferroptosis
```

## Multi-group GSEA

Multi-group GSEA requires a named list of ranked gene vectors:

```{r multi-group-gsea}
data("demo_rank_apop_list", package = "CDEnrich")

multi_gsea <- celldeath_gsea_multiple(
  geneList_list = demo_rank_apop_list,
  pvalueCutoff = 1
)

multi_gsea
```

A GSEA curve can be generated for the representative apoptosis pathway:

```{r multi-group-gsea-plot}
curves <- plot_death_gsea_curve_multiple(
  multi_gsea,
  pathway = apoptosis_id
)

curves$Control
curves$Treatment
```

The `pvalueCutoff = 1` setting retains results for all groups, including
groups without a strong enrichment signal, which facilitates comparison
across groups.

## ssGSEA and Differential Analysis

ssGSEA calculates a pathway score for each sample:

```{r ssgsea}
data("demo_expr", package = "CDEnrich")
data("demo_group", package = "CDEnrich")

ssgsea_result <- celldeath_ssgsea(
  expr = demo_expr,
  group = demo_group
)

ssgsea_result
```

A heatmap can be used to visualize pathway scores across samples:

```{r ssgsea-heatmap}
plot_death_ssgsea_heatmap(
  ssgsea_result
)
```

Differential analysis can be performed when group labels are available:

```{r ssgsea-differential}
ssgsea_difference <- celldeath_ssgsea_diff(
  ssgsea_result
)

head(ssgsea_difference)
```

For two groups, the function uses a Wilcoxon test. For three or more
groups, it uses a Kruskal-Wallis test.

Group-level scores can also be displayed with a boxplot:

```{r ssgsea-boxplot}
plot_death_ssgsea_boxplot(
  ssgsea_result
)
```

## Exporting Results

Results can be exported using the `savefile`, `filename`, or
`filename_prefix` arguments. The following examples are shown but not
executed during vignette construction:

```{r export-ora, eval=FALSE}
celldeath_enrich(
  deg = demo_deg,
  universe = demo_all_genes,
  savefile = TRUE,
  filename = "single_group_ORA_result.csv"
)
```

```{r export-gsea, eval=FALSE}
celldeath_gsea(
  geneList = demo_rank_apop,
  savefile = TRUE,
  filename = "single_group_GSEA_result.csv"
)
```

```{r export-multi-group-ora, eval=FALSE}
celldeath_compare_enrich(
  deg_list_list = demo_deg_list,
  universe = demo_all_genes,
  savefile = TRUE,
  filename = "multi_group_ORA_result.csv"
)
```

```{r export-multi-group-gsea, eval=FALSE}
celldeath_gsea_multiple(
  geneList_list = demo_rank_apop_list,
  savefile = TRUE
)
```

```{r export-ssgsea, eval=FALSE}
celldeath_ssgsea(
  expr = demo_expr,
  group = demo_group,
  savefile = TRUE,
  filename = "ssGSEA_score_matrix.csv"
)

celldeath_ssgsea_diff(
  ssgsea_result,
  savefile = TRUE,
  filename = "ssGSEA_differential_results.csv"
)
```

# Data Sources and Limitations

The curated gene sets were compiled from published literature and public
annotations. PMID-supported gene sets are linked to their corresponding
publications when PMID information is available. Some gene sets are
based primarily on GeneCards annotations and should be interpreted as
exploratory resources.

Enrichment results depend on the input genes, ranking statistic,
background gene set, pathway definition, and statistical thresholds.
They should be interpreted together with the experimental design and the
original publications. Enrichment does not by itself establish a causal
relationship or absolute pathway activity.

# Session Information

```{r session-info}
sessionInfo()
```
